Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments

### 2016-10-02 papers/2016-Special-Issue/v-special-issue-2016.tex
master now has the tex for the special issue submission -- but not the R -- for the R you have to git checkout special-issue

\title{Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments}
\date{\today}
\authornote{latex compiled \today ~\currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language use is vague, but the causes of this phenomenon are not well understood. Consequently, it is difficult for the designers of a Natural Language Generation (NLG) system to know when to let the system generate vague expressions. The present paper is an attempt to find out what benefits vagueness might have for readers.% , by means of a series of carefully controlled psycholinguistic experiments that focus on the use of vagueness in referring expressions. The work is motivated by NLG but has implications for linguistics, philosophical logic and game theory. 
%
This article reports on a series of experiments that aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. We argue that the evidence so far supports a view where the benefits that vague terms exert are due to other influences, rather than to vagueness itself. These factors include: low granularity; the use of evaluative words; the avoidance of overtly numerical words; the existence of comparison strategies; and, lastly and more tentatively, a phenomenon that we call range reduction. The paper concludes with a brief discussion of the implications for practical NLG.}

@@

NLG systems transform data and formulas into language \cite<e.g.,>[]{reiter2000building}. NLG systems routinely make decisions between different formulations of the same information. For example, if the temperature is 27.2 degrees Celsius, this could be expressed as ``27.2 degrees'', ``approximately 27 degrees', ``above 25 degrees", or ``warm'', and the system must decide between these.  The problem is especially important for NLG systems that take numbers as input, for example in the generation of textual weather reports from numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurement such as oxygen saturation, heart rhythm, etc. \cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. Such systems are often forced to make decisions concerning the level of precision in the utterances that they generate on the basis of little more than intuition. Even when NLG systems are designed to mimic human language use \cite<e.g.,>{konstas2013global} there is no guarantee that these decisions taken by these systems benefit readers. A better understanding of the benefits (for readers and hearers) of different precision levels would allow these systems to become more useful. The present article investigates the benefits,  or de-benefits, of vagueness.

@@

Language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.} In most academic use though, the word `vagueness' has a specific meaning. Keefe and Smith, for example, state ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \cite[p.\ 4]{keefe1997vagueness}, also \citeA{EgreKlinedinst}).  The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \cite[p.\ 1]{lipmanvague}. A typical example is the word ``tall'', as applied to people for example, because here is no precise, known height which separates those who are tall from those who are not. The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

@@

Lipman argued that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be. %, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.
Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, van2010vagueness}. Prominent among these answers is the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this as the \emph{cost reduction} hypothesis. %The idea of vagueness as cost reduction can take various shapes, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). 
%This hypothesis predicts that when a vague expression can perform the same communicative task as a precise one, then, the things being equal, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

@@

\subsection{General Methodology}\label{method}
The experiments reported in the present paper put the cost reduction hypothesis to the test. 
The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. 
Like Lipman, we focus on situations where numerical information is used in order to identify a referent. 
Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  
In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

In using references to quantities to test the cost reduction hypothesis we are only testing one aspect of vagueness in a particular context. 
This limits the applicability of our results. 
However, it has the advantage that it enables us to explore the costs and benefits of vagueness more thoroughly. 
Since one prevalent view of vagueness is that a vague expression is never preferable to a crisp equivalent, a demonstration of a benefit for vagueness in any context would advance the discussion.

@@

% MG General methodology moved to here: per-experiment method sections only talk about deviations from this general method
All the experiments shared the following properties: 
Stimuli were created using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
The position of the dots was randomised per-trial.
The order in which trials were presented was randomised per-participant. 
There were 256 trials, presented in 4 blocks of 64 each, between which the participant could rest. 
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants and recorded responses. 
Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
The experiment was conducted in a quiet room. 
Participants were asked to respond as quickly as possible while avoiding errors.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.
All $p$ values reported for linear models were calculated using the R package \emph{lmerTest} \cite{lmerTest}.

@@

When the distance grows between two numbers, they become more easily distinguishable from each other: the \emph{numerical distance effect}, which has been shown for comparing the cardinality of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene1996}. 
We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral. 
% MG added/moved out of expt 1
When two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. 
We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

% KvD added the following from a previous version of the paper:
There is abundant evidence \cite<e.g.,>{trick1994small} that very small (i.e., \emph{subitizable}) quantities are recognised and processed by a distinct psychological mechanism that differs from that used to process larger quantities. We performed a pilot experiment \cite{our-earlier-paper} in which we were able to confirm this finding in the experimental settings on which we are focussing in this paper. As one might expect, we found that, when participants were confronted with % MG changed the next line and added the footnote to remove the doubts we had about whether a dot array must have regular layout, or can have irregular layout - I'm still not sure this is a good description though
a stimulus consisting of two squares containing different numbers of dots\footnote{such a stimulus is referred to hereafter as a stimulus consisting of a set of some number of dot arrays, where the number of dots in each dot array varies and is referred to as its cardinality, and the physical arrangement of dots in each dot array is irregular}, 
instructions of the form \emph{Choose the square with \emph{n} dots} led to consistently faster response times than instructions of the form \emph{Choose the square with many/few dots} when $2 \leq n \leq 5$; the converse was true for $n>5$. We henceforth focussed our studies on non-subitizable numbers, because it is there that vagueness is expected to have benefits.

@@

\section{Experiment One}

\subsection{Introduction}

@@

\subsection{Hypothesis}
\begin{APAenumerate}
	\item [(H1)] A main effect advantage for vagueness: vague instructions impose a lower cognitive load for the comprehender than crisp alternatives.
\end{APAenumerate}

\subsection{Method}

@@

\begin{figure}[htbp]
\fitfigure{images/plotRTandErrorRates-1}
\caption{Experiment 1 results: response times and error rates.}
\label{resultse1}
\end{figure}

\subsubsection{Response times}
Response times (RTs) for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse1}).
A linear mixed model of RT was built, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants.

\begin{APAenumerate}
	\item [(H1$_{rt}$)] RTs were faster for vague instructions than for crisp instructions 
	($\beta=-0.11$, $se=.02$, $t=4.7$, $p<.001$).  % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}

\subsubsection{Error rates}
Error rate data were analysed using a generalized logit mixed model \cite{jaeger2008categorical}, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants.

\begin{APAenumerate}
	\item [(H1$_{er}$)] There was a statistically non-significant effect of vagueness on error rates: the vague instructions tended to have a lower error rate than the than the crisp instructions 
	($\beta=-.22$, $se=.18$, $t=-1.2$, $p=.230$).  % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}

@@

As expected, responses were faster and more accurate for vague than crisp instructions. Post hoc, we formulated some further hypotheses:

\begin{APAenumerate}
	\item [(H2)] A main effect advantage of increasing numerical distance: the task will become easier as the numerical distance increases, because the two arrays are then more discriminable.
	\item [(H3)] An interaction between vagueness and numerical distance: i.e., any facilitation for vagueness should be greater at smaller numerical distances than at larger numerical distances.
\end{APAenumerate}

@@

\begin{APAenumerate}
	\item [(H2$_{rt}$)] RTs increased with numerical distance 
	($\beta=-0.18$, $se=0.01$, $t=-11.6$, $p<0.0005$). % MG added the values for these stats on 25 Sept 2016
	Since discriminability of the arrays is easier for larger numerical distances, discriminability probably underlies this effect. 
	\item [(H3$_{rt}$)] Numerical distance and vagueness interacted significantly: essentially there were diminishing returns for vagueness as numerical distance increased: 
	($\beta=0.07$, $se=0.01$, $t=5.9$, $p<0.0005$). % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}

Looking at error rates, the two post-hoc hypotheses were supported:

\begin{APAenumerate}
	\item [(H2$_{er}$)] Error rates decreased as numerical distance increased: 
	($\beta=-0.78$, $se=0.08$, $z=-9.6$, $p<0.0005$).% MG added the values for these stats on 25 Sept 2016
	\item [(H3$_{er}$)] Numerical distance and vagueness interacted significantly: essentially there were diminishing returns for vagueness as numerical distance increased until the biggest numerical distance when vagueness was disadvantageous: 
	($\beta=0.80$, $se=0.15$, $z=5.27$, $p<0.0005$). % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}

So far, these results are in line with the idea of cost reduction. Cost reduction explains the vagueness advantage by claiming that the vague referring expressions place less cognitive load on the comprehender than the crisp referring expressions. It explains the diminishing returns for vagueness in more-discriminable stimuli (i.e., the vagueness by numerical distance interaction) by claiming that load is low in both conditions for the easily-discriminable stimuli, and that therefore there is no extra benefit to be had from vagueness in the easily-discriminable stimuli.

@@

\section{Experiment 2}

\subsection{Introduction} % E2

@@

To find out what happens when words are used in a context where their potential for vagueness comes to the fore,
Experiment 2 used three arrays so that the vague quantifiers always had more than one possible referent, and used indefinite articles in the vague instructions to avoid the impression that only one response counted as correct, and was carried out without error feedback.
An indication that the potential for vagueness was realised in Experiment 2 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 2, an item was a referring expression instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays had the following properties: it comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more.

Examples of crisp and vague versions of the numerical and verbal instructions follow: the examples assume the array (6,15,24) and reference to the smaller number of dots, such that 6 was classified as the expected response; 15 was classified as the borderline response; and 24 was classified as the incorrect response. In the {\em vague numerical} condition we used \emph{Choose a square with about 10 dots}. None of the squares contained 10 dots. 10 is slightly closer to 6 than to 15, justifying 6 as the best response and 15 as the borderline response. In the {\em vague verbal} condition we used \emph{Choose a square with few dots}. In the {\em crisp numerical} condition we used \emph{Choose the square with 6 dots}, and one square always did contain the number mentioned. For {\em crisp verbal}, we used \emph{Choose the square with the fewest dots}.

@@

\subsection{Hypotheses} % E2
\begin{APAenumerate}
	\item [(H1)] A main effect RT advantage for vagueness.
	\item [(H2)] An RT advantage for vagueness both in \emph{numeric} and in \emph{verbal} instructions.
	\item [(H3)] No large main effect of instruction format (of numeric versus verbal) since we hypothesised that vagueness rather than instruction format drove the effect in experiment 1.
	\item [(H4)] On the basis of Experiment 1, we would expect faster responses for stimuli with more discriminable arrays.(i.e., an effect of item).
	\item [(H5)] Participants should make more borderline case choices for vague than crisp instructions.
\end{APAenumerate}

\subsection{Method} % E2
% Participants were aged between 18 and 45 with a median age of 28. 
We manipulated as independent variables vagueness and instruction format, yielding four conditions, \emph{vague numeric}; \emph{vague verbal}; \emph{crisp numeric}; \emph{crisp verbal}.
We measured two dependent variables: response time; and the probability of a participant choosing the borderline case.
On each trial, first the referring expression that constituted the instruction for that trial was displayed. 
Participants then pressed a key to indicate that they had read the instruction. 
After 1000 ms, the arrays were presented, while preserving the text of the referring expression. 
The response time dependent variable was measured from the presentation of the arrays, until the keypress indicating the participant's choice, which was also recorded. 
The trial would timeout after 60 seconds if there was no response.
In this experiment, no feedback was given. 
This was because, in the vague conditions, we did not regard any response as `correct' or `incorrect', but instead as `borderline response', or `not borderline response', and we did not want to draw participants' attention to this distinction explicitly. 
We simply recorded whether the participant chose the best referent, the borderline case or the poorest referent, and how long it took the participant to respond.

@@

\subsubsection{Response times} % E2
% MG added the values for these stats on 26 Sept 2016
Means for response times and proportion of borderline responses are given in Fig. (\ref{resultse2}). 
Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. 
A linear mixed model was constructed for the (logged) response times, 
with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects.
\begin{APAenumerate}
	\item [(H1)] The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06$, $se=.01$, $t=4.6$, $p<0.0001$) 
	\item [(H2)] In focussed comparisons, vagueness was significantly disadvantageous in the \emph{numeric} conditions ($\beta=0.09$, $se=0.02$, $t=4.5$, $p<0.0001$), and non-significantly disadvantageous in the \emph{verbal} conditions ($\beta=0.02$, $se=0.02$, $t=1.5$, $p=0.1550$), offering partial evidence against hypothesis 2. The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to a significant overall interaction effect between vagueness and instruction format ($\beta=-0.13$, $se=0.02$, $t=-6.6$, $p=0.0169$).
	\item [(H3)] There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1, but suggesting that instruction format rather than vagueness drove the effect we observed in experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; ($\beta=-0.36$, $se=0.07$, $t=-5.1$, $p<0.0001$).
	\item [(H4)] There was a significant main effect of item (i.e., of which triple of numbers of dots were used in the stimulus): ($\beta=0.12$, $se=0.02$, $t=7.1$, $p<0.0001$). This effect seems likely to be due to the very fast responses for stimuli using arrangements of the smallest numbers of dots (6,15,24), which also had the largest difference in ratio of smallest number to largest number in the stimulus, suggesting that stimuli using these numbers of dots may have been particularly discriminable for participants.
\end{APAenumerate}

>>>>

% MG change p's to 3 dp

\title{Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments}
\date{\today}
%\authornote{latex compiled \today ~\currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language use is vague, but the causes of this phenomenon are not well understood. Consequently, it is difficult for the designers of a Natural Language Generation (NLG) system to know when to let the system generate vague expressions. The present paper is an attempt to find out what benefits vagueness might have for readers. % , by means of a series of carefully controlled psycholinguistic experiments that focus on the use of vagueness in referring expressions. The work is motivated by NLG but has implications for linguistics, philosophical logic and game theory. 
%
We report on a series of experiments that aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. We argue that the evidence so far supports a view where the benefits that vague terms exert are due to other influences, rather than to vagueness itself. These factors include: low granularity; the use of evaluative words; the avoidance of overtly numerical words; the existence of comparison strategies; and, lastly and more tentatively, a phenomenon that we call range reduction. The paper concludes with a discussion of the implications for practical NLG.}

@@

NLG systems transform data and formulas into language \cite<e.g.,>[]{reiter2000building}. NLG systems routinely make decisions between different formulations of the same information. For example, if the temperature is 27.2 degrees Celsius, this could be expressed as ``27.2 degrees'', ``approximately 27 degrees', ``above 25 degrees", or ``warm'', and the system must decide between these.  The problem is especially important for NLG systems that take numbers as input, for example in the generation of textual weather reports from numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurements % such as oxygen saturation, heart rhythm, etc.
\cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. Such systems are often forced to make decisions concerning the level of precision in the utterances that they generate on the basis of little more than intuition. Even when NLG systems are designed to mimic human language use \cite<e.g.,>{konstas2013global} there is no guarantee that these decisions taken by these systems benefit readers. A better understanding of the benefits (for readers and hearers) of different precision levels would allow these systems to become more useful. The present article investigates the benefits,  or de-benefits, of vagueness.

@@

%Language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.}

In most academic use, the word `vagueness' has a specific meaning. Keefe and Smith, for example, state ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \cite[p.\ 4]{keefe1997vagueness}, also \citeA{EgreKlinedinst}).  The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \cite[p.\ 1]{lipmanvague}. A typical example is the word ``tall'', as applied to people for example, because here is no precise, known height which separates those who are tall from those who are not. The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

@@

Lipman argued that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be. %, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.
Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, vanDeemterBook}. Prominent among these answers is the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this as the \emph{cost reduction} hypothesis. %The idea of vagueness as cost reduction can take various shapes, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). 
%This hypothesis predicts that when a vague expression can perform the same communicative task as a precise one, then, the things being equal, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

@@

\subsection{General Methodology}\label{method}
%The experiments reported in the present paper put the cost reduction hypothesis to the test. The question that we are trying to answer is whether

Are vague expressions processed more easily by readers than crisp ones? 
Like Lipman, we focus on situations where numerical information is used in order to identify a referent. 
Reference, in other words, will be the communicative task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  
% In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

In using references to quantities to test the cost reduction hypothesis we are only testing one aspect of vagueness in a particular context. 
This limits the direct applicability of our results. 
However, it has the advantage that it enables us to explore the costs and benefits of vagueness more thoroughly. 
Since one prevalent view of vagueness is that a vague expression is never preferable to a crisp equivalent, a demonstration of a benefit for vagueness in any context would advance the discussion.

@@

% MG General methodology moved to here: per-experiment method sections only talk about deviations from this general method
All the experiments shared the following properties: 
Stimuli were created using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
The position of the dots was randomised per-trial.
The order in which trials were presented was randomised per-participant. 
There were 256 trials, presented in 4 blocks of 64 each, between which the participant could rest. 
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants and recorded responses. 
Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
The experiment was conducted in a quiet room. 
Participants were asked to respond as quickly as possible while avoiding errors.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.
All $p$ values reported for linear models were calculated using the R package \emph{lmerTest} \cite{lmerTest}. The complete data and analysis of our first experiment are available from 
{\tt http://homepages.abdn.ac.uk/k.vdeemter/pages/vagueness-data-analysis/e1.html}, and analogous for experiments 2-4.\\[2ex]

@@

\noindent When the distance grows between two numbers, they become more easily distinguishable: the \emph{numerical distance effect} has been shown for comparing the cardinality of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene1996}. 
We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral. 
% MG added/moved out of expt 1
When two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. 
We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

% KvD added the following from a previous version of the paper:
There is abundant evidence \cite<e.g.,>{trick1994small} that very small (i.e., \emph{subitizable}) quantities are recognised and processed by a distinct psychological mechanism that differs from that used to process larger quantities. We performed a pilot experiment \cite{our-earlier-paper} in which we were able to confirm this finding in the experimental settings on which we are focussing in this paper. We found that, when participants were confronted with % MG changed the next line and added the footnote to remove the doubts we had about whether a dot array must have regular layout, or can have irregular layout - I'm still not sure this is a good description though
a stimulus consisting of two squares containing different numbers of dots\footnote{Such a stimulus is referred to hereafter as consisting of a set of {\em dot arrays}. The number of dots in an array is referred to as its cardinality. The physical arrangement of dots in each array is irregular.}, 
instructions of the form \emph{Choose the square with \emph{n} dots} led to consistently faster response times than instructions of the form \emph{Choose the square with many/few dots} when $2 \leq n \leq 5$; the converse was true for $n>5$. Given these findings, we henceforth focussed our studies on non-subitizable numbers, because it is there that vagueness is expected to have benefits.

@@

\section{Experiment 1: initial experiment}

@@

\subsection{Hypothesis}
\noindent The main aim of this experiment was to test the following hypothesis:
{\small
\begin{APAenumerate}
	\item [(H1)] There exists a main effect advantage for vagueness: vague instructions impose a lower cognitive load for the comprehender than crisp alternatives.
\end{APAenumerate}
}

\subsection{Experimental Method}

@@

\begin{figure}[htbp]
\fitfigure{images/e1RTandErrorRatesplot-1}
\caption{Experiment 1 results: response times and error rates.}
\label{resultse1}
\end{figure}

%\subsubsection{Response times}
{\bf Response times} (RTs) for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse1}).
A linear mixed model of RT was built, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants. Analysis confirms hypothesis H1:

{\small
\begin{APAenumerate}
	\item [(H1$_{rt}$)] RTs were faster for vague instructions than for crisp instructions 
	($\beta=-0.11$, $se=.02$, $t=4.7$, $p<.001$).  % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}
}

%\subsubsection{Error rates}
{\bf Error rate} data were analysed using a generalized logit mixed model \cite{jaeger2008categorical}, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants. We found:

{\small
\begin{APAenumerate}
	\item [(H1$_{er}$)] There was a statistically non-significant effect of vagueness on error rates: the vague instructions tended to have a lower error rate than the than the crisp instructions 
	($\beta=-.22$, $se=.18$, $t=-1.2$, $p=.230$).  % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}
}

@@

\noindent As expected, responses were faster and more accurate for vague than crisp instructions. Post hoc, we formulated some further hypotheses:

{\small
\begin{APAenumerate}
	\item [(H2)] A main effect advantage of increasing numerical distance: the task will become easier as the numerical distance increases, because the two arrays are then more discriminable.
	\item [(H3)] An interaction between vagueness and numerical distance: i.e., any facilitation for vagueness should be greater at smaller numerical distances than at larger numerical distances.
\end{APAenumerate}
}

@@

{\small
\begin{APAenumerate}
	\item [(H2$_{rt}$)] RTs increased with numerical distance 
	($\beta=-0.18$, $se=0.01$, $t=-11.6$, $p<0.0005$). % MG added the values for these stats on 25 Sept 2016
	Since discriminability of the arrays is easier for larger numerical distances, discriminability probably underlies this effect. 
	\item [(H3$_{rt}$)] Numerical distance and vagueness interacted significantly: essentially there were diminishing returns for vagueness as numerical distance increased: 
	($\beta=0.07$, $se=0.01$, $t=5.9$, $p<0.0005$). % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}
}

Looking at error rates, the two post-hoc hypotheses were supported as well:

{\small
\begin{APAenumerate}
	\item [(H2$_{er}$)] Error rates decreased as numerical distance increased: 
	($\beta=-0.78$, $se=0.08$, $z=-9.6$, $p<0.0005$).% MG added the values for these stats on 25 Sept 2016
	\item [(H3$_{er}$)] Numerical distance and vagueness interacted significantly: essentially there were diminishing returns for vagueness as numerical distance increased until the biggest numerical distance when vagueness was disadvantageous: 
	($\beta=0.80$, $se=0.15$, $z=5.27$, $p<0.0005$). % MG added the values for these stats on 25 Sept 2016
\end{APAenumerate}
}

So far, these results are in line with the idea of cost reduction. The idea of cost reduction explains the vagueness advantage by claiming that the vague referring expressions place less cognitive load on the comprehender than the crisp referring expressions. It explains the diminishing returns for vagueness in more-discriminable stimuli (i.e., the vagueness by numerical distance interaction) by claiming that load is low in both conditions for the easily-discriminable stimuli, and that therefore there is no extra benefit to be had from vagueness in the easily-discriminable stimuli.

@@

\section{Experiment 2: separating vagueness from instruction format}

@@

To find out what happens when words are used in a context where their potential for vagueness comes to the fore, Experiment 2 used three arrays (rather than two) so that the vague description had more than one possible referent, and used indefinite articles to avoid the impression that only one response counted as correct, and was carried out without error feedback.
An indication that the potential for vagueness was realised in Experiment 2 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 2, an item was a referring expression instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more.

Examples of crisp and vague versions of the numerical and verbal instructions follow: the examples assume the array (6,15,24) and reference to the smaller number of dots, such that 6 was classified as the expected response; 15 as the borderline response; and 24 as the incorrect response. In the {\em vague numerical} condition we used \emph{Choose a square with about 10 dots}. None of the squares contained 10 dots. 10 is slightly closer to 6 than to 15, justifying 6 as the best response and 15 as the borderline response. In the {\em vague verbal} condition we used \emph{Choose a square with few dots}. In the {\em crisp numerical} condition we used \emph{Choose the square with 6 dots}, and one square always did contain the number mentioned. For {\em crisp verbal}, we used \emph{Choose the square with the fewest dots}.

@@

\subsection{Hypotheses} % E2
\noindent We formulated the following hypotheses for Experiment 2:

{\small
\begin{APAenumerate}
	\item [(H1)] There exists a main effect RT advantage for vagueness.
	\item [(H2)] An RT advantage for vagueness both in \emph{numeric} and in \emph{verbal} instructions.
	\item [(H3)] No large main effect of instruction format (i.e., numeric versus verbal) since we hypothesised that vagueness rather than instruction format drove the effect in Experiment 1.
	\item [(H4)] On the basis of Experiment 1, we expect faster responses for stimuli with more discriminable arrays.(i.e., an effect of item).
	\item [(H5)] Participants should make more borderline case choices for vague than crisp instructions.
\end{APAenumerate}
}

\subsection{Experimental Method} % E2
% Participants were aged between 18 and 45 with a median age of 28. 
We manipulated as independent variables vagueness and instruction format (i.e., numeric versus verbal instructions), yielding four conditions, \emph{vague numeric}; \emph{vague verbal}; \emph{crisp numeric}; \emph{crisp verbal}.
We measured two dependent variables: response time; and the probability of a participant choosing the borderline case.
On each trial, first the referring expression that constituted the instruction for that trial was displayed. 
Participants then pressed a key to indicate that they had read the instruction. 
After 1000 ms, the arrays were presented, while preserving the text of the referring expression. 
The response time dependent variable was measured from the presentation of the arrays, until the keypress indicating the participant's choice, which was also recorded. 
The trial would timeout after 60 seconds if there was no response.
In this experiment, no feedback was given. 
This was because, in the vague conditions, we did not regard any response as `correct' or `incorrect', but instead as `borderline response', or `not borderline response', and we did not want to draw participants' attention to this distinction explicitly. 
We simply recorded whether the participant chose the best referent, the borderline case or the poorest referent, and how long it took the participant to respond.

@@

%\subsubsection{Response times} % E2
% MG added the values for these stats on 26 Sept 2016
Means for response times and proportion of borderline responses are given in Fig. (\ref{resultse2}). 
Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. 
A linear mixed model was constructed for the (logged) response times, 
with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. We found:
{\small
\begin{APAenumerate}
	\item [(H1)] The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against Hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06$, $se=.01$, $t=4.6$, $p<0.0001$). 
	\item [(H2)] In focussed comparisons, vagueness was significantly disadvantageous in the \emph{numeric} conditions ($\beta=0.09$, $se=0.02$, $t=4.5$, $p<0.0001$), and non-significantly disadvantageous in the \emph{verbal} conditions ($\beta=0.02$, $se=0.02$, $t=1.5$, $p=0.1550$), offering partial evidence against Hypothesis 2. The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to a significant overall interaction effect between vagueness and instruction format ($\beta=-0.13$, $se=0.02$, $t=-6.6$, $p=0.0169$).
	\item [(H3)] There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1, but suggesting that instruction format rather than vagueness drove the effect we observed in Experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; ($\beta=-0.36$, $se=0.07$, $t=-5.1$, $p<0.0001$).
	\item [(H4)] There was a significant main effect of item (i.e., of which triple of numbers of dots were used in the stimulus): ($\beta=0.12$, $se=0.02$, $t=7.1$, $p<0.0001$). This effect seems likely to be due to the very fast responses for stimuli using arrangements of the smallest numbers of dots (6,15,24), which also had the largest difference in ratio of smallest number to largest number in the stimulus, suggesting that stimuli using these numbers of dots may have been particularly discriminable for participants.
\end{APAenumerate}
}
